---
title: "BioLizardStyleR In Action"
author: "Robbe Neirynck"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this guide, we'll quickly walk you through the installation of the BioLizardStyleR package and provide an overview of the functions
using a practical example.

The BioLizardStyleR package helps you make ggplot2 plots in the BioLizard style. It includes an integrated font installer, a dedicated ggplot2 theme, custom color functions, and a tool to add a BioLizard footer and export the plot.

#1. Package Installation
To install the `BioLizardStyleR` package directly from GitHub, you will need to utilize the `devtools` package.
```{r}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("lizard-bio/nature-grade-visualization-playground", subdir="BioLizardStyleR",
                         auth_token = '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')

library(BioLizardStyleR)
```
NOTE: Since package is private you will need a Personal Access Token.
-Go to your GitHub settings (click on your profile picture in the top-right corner and select "Settings").
-On the left sidebar, select "Developer settings".
-Click on "Personal access tokens".
-Click "Generate new token".
-Give your token a name, and select the repo scope, which allows access to private repositories. Then generate the token.
-Copy the token (make sure to save it somewhere, as you won't be able to see it again).

#2. Font Installation
Avenir, the signature font of BioLizard, is incorporated within the package. You can install this font using the `install_biolizard_fonts()` function. Note that this installation is a one-time process and doesn't need to be repeated in subsequent R sessions.
```{r}
install_biolizard_fonts()
```

#3. Creating your BioLizard-Styled ggplot

Transforming your ggplot to match the BioLizard style involves a three-step process. I'll illustrate this using the renowned `mtcars` dataset.
Let's start of with creating a basic ggplot.
```{r}
library(ggplot2)

basic1 <- ggplot(mtcars, aes(x=factor(cyl))) +
            geom_bar() + 
            xlab('Number of Cylinders') + 
            ylab('Count') + 
            ggtitle('Number of Cars by Cylinders')

basic2 <- ggplot(mtcars, aes(x=rownames(mtcars), y=hp, fill=hp)) + 
            geom_bar(stat="identity") +
            scale_fill_gradient(low="blue", high="red") +
            xlab('Car Model') + 
            ylab('Horsepower') + 
            ggtitle('Horsepower of Cars with a Sequential Color Palette')



```


