---
title: "BioLizardStyleR In Action"
author: "BioLizard"
date: "07-03-2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction
In this guide, we'll walk you through the installation of the `BioLizardStyleR` package and provide a brief overview of its functions using a practical example. The `BioLizardStyleR` package helps you make ggplot2 plots in the BioLizard style. It includes an integrated font installer, a dedicated ggplot2 theme, custom color functions, and a tool to add a BioLizard footer and export the plot.


# 1. Installation
To install the `BioLizardStyleR` package directly from GitHub, you will need to utilize the `devtools` package. The package requires our signature font "Lato". If the font is not yet installed, installation  will be attempted automatically when you load the library.

```{r}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("lizard-bio/nature-grade-visualization-playground", subdir="BioLizardStyleR")

library(BioLizardStyleR)
```

If automatic font installation fails on your system, you can install the font manually using
the following steps:

1. **Download Fonts**: Download the font files from the
GitHub repository [Nature Grade Visualization
Playground](https://github.com/lizard-bio/nature-grade-visualization-playground/tree/main).
2. **Install Fonts**: Install the downloaded fonts on your system with your favorite font manager. Double-clicking the .ttf file will already open your font manager in many cases.
<!-- 3. **Register Fonts**: On other systems, once the installation is done, you can run -->
<!-- `install_biolizard_fonts()` **without any arguments** to register the font with -->
<!-- the extrafont package. If you encounter issues due to a pre-existing -->
<!-- installation of the extrafont package and need to clear the database, use the -->
<!-- argument `clearDatabase=TRUE`. This will clear the cache of the extrafontdb -->
<!-- package to reset its database. Note that this installation is a one-time process -->
<!-- and normally doesn't need to be repeated in subsequent R sessions. -->
```
install_biolizard_fonts()
```

# 2. Help & Documentation

The documentation is hosted on github pages: [BioLizardStyleR](https://lizard-bio.github.io/nature-grade-visualization-playground/BioLizardStyleR/)

Alternatively, if you installed the package using `build_vignettes = TRUE`, you can check out the "BioLizardStyleR" vignette, which contains many examples of plots you might want to make in the future, including (interactive) plots for a simple RNA-seq case-control analysis.

```{r}
vignette("BioLizardStyleR")
```

Help files for the individual functions can be called as usual in R:

```{r}
?biolizard_pal_qualitative
```


# 3. Crafting a BioLizard-Styled ggplot

Transforming your ggplot to match the BioLizard style involves a three-step process. Let's illustrate this on a basic ggplot.

```{r , fig.width=6}
library(ggplot2)
data("mtcars")
mtcars$gear <- factor(mtcars$gear, levels = c(3, 4, 5), ordered = TRUE)

testplot <- ggplot(data = mtcars, aes(x = hp, y = mpg)) + 
              geom_point(aes(color = gear),size=3) +  
              labs(title = "Miles per Gallon vs. Horsepower",  
                   x = "Horsepower",  
                   y = "Miles per Gallon",  
                   color = "Gears")
testplot
```

## 3.1 Applying the Biolizard theme

In the initial step, we implement the biolizard theme through the `lizard_style()` function. This not only modifies various theme settings but also adjusts the font. An important note here is that the lizard theme can only act as a starting template! Depending on your specific plot, title length, variables in use, and other factors, you might need to make further theme adjustments.

```{r , fig.width=6}
testplot <- testplot + lizard_style()
testplot
```

**Tip:** you can use `theme_set(lizard_style())` at the beginning of your script or report to set the theme to lizard_style() in all plots, without having to add `+ lizard_style()` to each plot individually. Te reset the theme to the ggplot2 default, you can call `reset_theme_settings()`.

## 3.2 Adding a BioLizard color palette

Once you've styled your plot, you can enhance it with a 'BioLizard' color palette. You have the option to choose from qualitative, sequential, or divergent palettes tailored for either discrete or continuous variables. These palettes incorporate the distinct hues of the BioLizard house colors, while ensuring they are color-blind friendly and perceptually uniform. For a deeper understanding of each palette, it's recommended to consult the documentation for the respective function.

The coloring functions within the package adhere to a flexible structure: scale_color/fill_biolizard. The type (discrete/continuous) and the scheme (qualitative/sequential/divergent) must be determined as function parameters.

```{r , fig.width=6}
data("mtcars")
mtcars$gear <- as.factor(mtcars$gear)

# Base plot
testplot <- ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point(aes(color = gear), size=3) +
  labs(
    title = "Miles per Gallon vs. Horsepower",
    x = "Horsepower",
    y = "Miles per Gallon",
    color = "Gears"
  )

# Applying the qualitative color scale
testplot_qualitative <- testplot + scale_color_biolizard(type = "discrete", scheme = "qualitative") + lizard_style()
testplot_qualitative

# Applying the sequential color scale (assuming you have a different dataset or aesthetic)
testplot_sequential <- testplot + scale_color_biolizard(type = "discrete", scheme = "sequential") + lizard_style()
testplot_sequential

# Applying the divergent color scale (assuming you have a different dataset or aesthetic)
testplot_divergent <- testplot + scale_color_biolizard(type = "discrete", scheme = "divergent") + lizard_style()
testplot_divergent
```

It is possible to reverse the color scales by setting `reverse = TRUE`

```{r , fig.width=6}
testplot_divergent <- testplot + scale_color_biolizard(type = "discrete", scheme = "divergent", reverse = TRUE) + lizard_style()
testplot_divergent
```

The discrete color palettes can also be called directly, in case you would like to use them outside ggplot. The three base colors can also be used directly as `blz_green`, `blz_blue` and `blz_yellow`.

```{r, fig.show='hold', fig.height=4, fig.width=3.3}
#use color palette
mycolors <- biolizard_pal_sequential(3, reverse = TRUE)
gearcolors <- mycolors[as.factor(mtcars$gear)]

plot(x = mtcars$hp, y = mtcars$mpg, pch = 19, col = gearcolors, family = "Lato")
legend("topright", legend = paste("Gears", 3:5), col = mycolors, pch = 19, bty = "n")

#use BLZ base colors
mycolors <- c(blz_green, blz_blue, blz_yellow)
gearcolors <- mycolors[as.factor(mtcars$gear)]

plot(x = mtcars$hp, y = mtcars$mpg, pch = 19, col = gearcolors, family = "Lato")
legend("topright", legend = paste("Gears", 3:5), col = mycolors, pch = 19, bty = "n")

```


## 3.3 Adding a BioLizard footer and exporting the plot.

For a unique touch on your visualizations, integrate a BioLizard footer using the **finalise_lizardplot()** function. This function not only positions the logo neatly at the bottom right but also provides an option to specify your data source directly in the footer, offering an alternative to the caption section of ggplot. When ready to export, either assign a name to your plot using `output_name` or specify a directory with `save_filepath`. The `device` argument allows to choose from the file types "png" (default), "pdf", "svg", or "ps".

```{r}
finalise_lizardplot(testplot + scale_color_biolizard(type = "discrete", scheme = "qualitative") + lizard_style(),
                    source='Source: The mtcars dataset')
```

## 3.4 Reset ggplot2 defaults

Note that calling `lizard_style()` will overwrite the default colors for points, lines, boxplots, areas, etc.. . To reset them to the ggplot2 defaults, you can either restart the R session or use `reset_colors()`.

In the plot below, the points are shown in <span style="color: #01a086;">BioLizard green</span>. Using `reset_colors()` will revert these changes:

```{r, fig.show='hold'}
p <- ggplot(mtcars, aes(mpg, disp)) + geom_point()
p

reset_colors()
p

```

# 4. Some examples for fun

## Simple box plot

```{r}
ggplot(data = mtcars, aes(x = as.factor(gear), y = mpg)) +
  geom_boxplot() +
  labs(
    title = "Miles per Gallon vs. Gears",
    x = "Gears",
    y = "Miles per Gallon"
  ) +
  lizard_style()

ggplot(data = mtcars, aes(x = as.factor(gear), y = mpg)) +
  geom_violin() +
  labs(
    title = "Miles per Gallon vs. Gears",
    x = "Gears",
    y = "Miles per Gallon"
  ) +
  lizard_style()
```

## Scatterplot

```{r}

ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Miles per Gallon vs. horsepower",
    x = "Horsepower",
    y = "Miles per Gallon",
    color = "Miles per Gallon",
    fill = "Miles per Gallon"
  )

```

## Facets
Faceting is a great way to lay out related plots side-by-side, with the axes either on the same scale or on different scales.

```{r,  fig.show='hold', fig.width=3.3}
# use facet_wrap to stratify plots by one variable
ggplot(mtcars, aes(x=mpg, y=hp)) +
  geom_point() +
  facet_wrap(vars(gear)) +
  lizard_style() +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))


# use facet_grid for two variables
ggplot(mtcars, aes(x=mpg, y=hp)) +
  geom_point() +
  facet_grid(rows = vars(gear), cols = vars(carb)) +
  lizard_style() +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))

```

Note that the minimal style makes it difficult to distinguish the different blocks from each other. Readability of the figure is improved by plotting the axes for each plot, but they are removed by `facet_grid` and `facet_wrap` when the plots are on the same scale. The similar functions `facet_rep_grid` and `facet_rep_wrap` from the lemon package are more suitable with the lizard style, since they will show all axes.

```{r,  fig.show='hold', fig.width=3.3}
if(!require("lemon", quietly = TRUE)){
  BiocManager::install("lemon")
}

library(lemon)

# use facet_wrap to stratify plots by one variable
ggplot(mtcars, aes(x=mpg, y=hp)) +
  geom_point() +
  facet_rep_wrap(vars(gear)) +
  lizard_style() +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))


# use facet_grid for two variables
ggplot(mtcars, aes(x=mpg, y=hp)) +
  geom_point() +
  facet_rep_grid(rows = vars(gear), cols = vars(carb)) +
  lizard_style() +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))

```
## Maps

note: to install 'sf' on ubuntu I had to install libudunits2-dev and libgdal-dev:

- debian/ubuntu: `sudo apt-get update && sudo apt-get install libudunits2-dev libgdal-dev`
- tophat/fedora: `dnf install udunits2-devel libgdal-devel`
- OSX: `brew install udunits` & `brew install gdal`

```{r}
if (!requireNamespace("sf", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("sf")
}
if (!requireNamespace("rnaturalearth", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("rnaturalearth")
}
if (!requireNamespace("rnaturalearthdata", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("rnaturalearthdata")
}

library("sf")
library("rnaturalearth")
library("rnaturalearthdata")

world <- ne_countries(scale = "medium", returnclass = "sf")

BLZ_offices <- c("Belgium", "Netherlands", "Switzerland", "United States of America")
world$BLZ_office <- ifelse(world$name_en %in% BLZ_offices, "yes", "no")
world$BLZ_office <- factor(world$BLZ_office, levels = c("yes", "no"))

ggplot(data = world, aes(fill = BLZ_office)) +
  geom_sf() +
  scale_fill_biolizard(type = "discrete", scheme = "qualitative") +
  lizard_style()

```

```{r}
# coloring the ocean, as in the map on our website
ggplot(data = world, aes(fill = BLZ_office)) +
  geom_sf() +
  scale_fill_manual(values = c("#e9b940", "#01a086")) +
  lizard_style() +
  theme(panel.background = element_rect(fill = "#1e2237"),
        panel.grid = element_blank())
```


```{r}
# europe only
eu_plot <- ggplot(data = dplyr::filter(world, continent == "Europe"), aes(fill = BLZ_office)) +
  geom_sf() +
  scale_fill_manual(values = c("#e9b940", "#01a086")) +
  lizard_style() +
  theme(panel.background = element_rect(fill = "#1e2237"),
        panel.grid = element_blank()) +
  coord_sf(xlim = c(-25,50), ylim = c(35,70), expand = FALSE)
eu_plot


finalise_lizardplot(eu_plot,source="BioLizard offices in Europe")
```

## Treemap plots

Treemap plots are a powerful way of visualizing hierarchical data such as GO terms, for example. The size of the rectangles corresponds to the value you want to show (e.g. p value), and related rectangles are grouped together. This example makes use of the Pokémon dataset from the [highcharter](https://cran.r-project.org/web/packages/highcharter/index.html) package and is based on https://yjunechoe.github.io/posts/2020-06-30-treemap-with-ggplot/.

The example below also illustrates how the `colorRampPalette` function can be used to expand a discrete color scale if the number of colors is not sufficient. Note that a discrete, qualitative color scale may not be very useful for a large number of colors, as it will be hard to distinguish the individual colors from one another.

The treemapify package allows plotting treemaps with ggplot2.

```{r , fig.width=6, fig.height=4}
if(!require("treemapify", quietly = TRUE)){
  BiocManager::install("treemapify")
}

if(!require("highcharter", quietly = TRUE)){
  install.packages("highcharter")
}

library(treemapify)
library(dplyr)

data("pokemon", package = "highcharter")
# Cleaning up data for a treemap
data <- pokemon |> 
  select(pokemon, type_1, type_2) |>
  mutate(type_2 = ifelse(is.na(type_2), paste("only", type_1), type_2)) |> 
  group_by(type_1, type_2) |>
  summarise(n = length(pokemon)) |> 
  ungroup()

head(data)

# expand the biolizard qualitative palette using colorRampPalette
ncols <- length(unique(data$type_1))
BLZ_qual <- colorRampPalette(biolizard_pal_qualitative(8))(ncols)

ggplot(data, aes(area = n, label = type_2, fill = type_1,
                subgroup = type_1)) +
  ggtitle("Number of pokémon stratified by type") +
  # 1. Draw type_2 borders and fill colors
  geom_treemap() +
  # 2. Draw type_1 borders
  geom_treemap_subgroup_border() +
  # 3. Print type_1 text
  geom_treemap_subgroup_text(place = "centre", grow = T, alpha = 0.5, colour = "black",
                             fontface = "italic", min.size = 0,
                             family = "Lato") +   #need to add "Lato" manually as this text is not accessed in theme()
  # 4. Print type_2 text
  geom_treemap_text(colour = "white", place = "topleft", reflow = T,
                    family = "Lato") +   #need to add "Lato" manually as this text is not accessed in theme()
  
  # 5. BLZ style
  scale_fill_manual(values = BLZ_qual) +
  lizard_style() +
  theme(legend.position = "none")

```

However, this is not ideal, since the font of the text still has to be defined manually, and thus the lizard_style() function does not work very well for these treemaps, even though they are ggplot2-based. Another option is to use the 'treemap' library, and move away from ggplot2 altogether, which generally leads to more aesthetically pleasing plots:

```{r, fig.width=6, fig.height=4}
if(!require("treemap", quietly = TRUE)){
  BiocManager::install("treemap")
}

library(treemap)

treemap(data, index=c("type_1", "type_2"), vSize="n", type="index",
        title="Number of pokémon stratified by type", 
        palette=BLZ_qual,  #set the BLZ color palette
        fontcolor.labels=c("#FFFFFFDD", "#00000080"), bg.labels=0, border.col="#00000080",
        fontfamily.title = "Lato", fontfamily.labels = "Lato", fontfamily.legend = "Lato")  #set the lato font

```

