---
title: "BioLizardStyleR In Action"
author: "Robbe Neirynck"
date: "17-08-2023"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Introduction
In this guide, we'll walk you through the installation of the `BioLizardStyleR` package and provide a brief overview of its functions using a practical example. The `BioLizardStyleR` package helps you make ggplot2 plots in the BioLizard style. It includes an integrated font installer, a dedicated ggplot2 theme, custom color functions, and a tool to add a BioLizard footer and export the plot.


# 1. Installation
To install the `BioLizardStyleR` package directly from GitHub, you will need to utilize the `devtools` package.
```{r}
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

devtools::install_github("lizard-bio/nature-grade-visualization-playground", subdir="BioLizardStyleR", ref = "aniko")

library(BioLizardStyleR)
```

# 2. Font Installation
The BioLizardStyleR package makes use of the 'Nunito Sans' font. This font resembles the 'Avenir', signature BioLizard font but can be found on Google Fonts under an open source license.

## Check Local Installation
First, check if you have the Nunito Sans font locally installed on your system. If so please delete this local installation.

## Download and Install from GitHub
If you don't have the Nunito Sans font locally installed, you can download and install it from GitHub.

1. **Download Fonts**: Download the two font files from the GitHub repository [Nature Grade Visualization Playground](https://github.com/lizard-bio/nature-grade-visualization-playground/tree/main).
2. **Install Fonts**: Install the downloaded fonts on your system.
3. **Complete**: Once the installation is done, you're all set to use the font.


Note that this installation is a one-time process and normally doesn't need to be repeated in subsequent R sessions. Running the following function **without any arguments** should install the font. If you encounter issues due to a pre-existing installation of the extrafont package and need to clear the database, use the argument clearDatabase=TRUE. This will reinstall the extrafontdb package, a companion to extrafont, to reset its database.
```{r}
install_biolizard_fonts()
```


# 3. Crafting a BioLizard-Styled ggplot

Transforming your ggplot to match the BioLizard style involves a three-step process. Let's illustrate this on a basic ggplots.

```{r}
library(ggplot2)
data("mtcars")
mtcars$gear <- factor(mtcars$gear, levels = c(3, 4, 5), ordered = TRUE)

testplot <- ggplot(data = mtcars, aes(x = hp, y = mpg)) + 
              geom_point(aes(color = gear),size=3) +  
              labs(title = "Miles per Gallon vs. Horsepower",  
                   x = "Horsepower",  
                   y = "Miles per Gallon",  
                   color = "Gears")
testplot
```

## 3.1 Applying the Biolizard theme

In the initial step, we implement the biolizard theme through the **lizard_style()* function. This not only modifies various theme settings but also adjusts the font. An important note here is that the lizard theme can only act as a starting template! Depending on your specific plot, title length, variables in use, and other factors, you might need to make further theme adjustments.
```{r}
testplot <- testplot + lizard_style()
testplot
```

## 3.2 Adding a BioLizard color palette

Once you've styled your plot, you can enhance it with a 'BioLizard' color palette. You have the option to choose from qualitative, sequential, or divergent palettes tailored for either discrete or continuous variables. These palettes incorporate the distinct hues of the BioLizard house colors, while ensuring they are color-blind friendly and perceptually uniform. For a deeper understanding of each palette, it's recommended to consult the documentation for the respective function.

The coloring functions within the package adhere to a flexible structure: **scale_<color/fill>_biolizard**. The type (discrete/continuous) and the scheme (qualitative/sequential/divergent) must be determined as function parameters!
```{r}
data("mtcars")
mtcars$gear <- as.factor(mtcars$gear)

# Base plot
testplot <- ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point(aes(color = gear), size=3) +
  labs(
    title = "Miles per Gallon vs. Horsepower",
    x = "Horsepower",
    y = "Miles per Gallon",
    color = "Gears"
  )

# Applying the qualitative color scale
testplot_qualitative <- testplot + scale_color_biolizard(type = "discrete", scheme = "qualitative") + lizard_style()
testplot_qualitative

# Applying the sequential color scale (assuming you have a different dataset or aesthetic)
testplot_sequential <- testplot + scale_color_biolizard(type = "discrete", scheme = "sequential") + lizard_style()
testplot_sequential

# Applying the divergent color scale (assuming you have a different dataset or aesthetic)
testplot_divergent <- testplot + scale_color_biolizard(type = "discrete", scheme = "divergent") + lizard_style()
testplot_divergent
```


## 3.3 Adding a BioLizard footer and exporting the plot.

For a unique touch on your visualizations, integrate a BioLizard footer using the **finalise_lizardplot()** function. This function not only positions the logo neatly at the bottom right but also provides an option to specify your data source directly in the footer, offering an alternative to the caption section of ggplot. When ready to export, either assign a name to your plot using `output_name` or specify a directory with `save_filepath`. The `device` argument allows to choose from the file types "png" (default), "pdf", "svg", or "ps".

```{r}
finalise_lizardplot(testplot + scale_color_biolizard(type = "discrete", scheme = "qualitative") + lizard_style(), source='Source: The mtcars dataset')
```

# 4. Some examples for fun

## Simple bar chart

```{r}
x = 1:10
y = 1:10

df <- data.frame(x=x, y=y)
ggplot(df, aes(x=x, y=y)) +
  geom_col() +
  lizard_style()
```

## Simple polygon plot

```{r}

ids <- factor(c("1.1", "2.1", "1.2", "2.2", "1.3", "2.3"))
values <- data.frame(
  id = ids,
  value = c(3, 3.1, 3.1, 3.2, 3.15, 3.5)
)
positions <- data.frame(
  id = rep(ids, each = 4),
  x = c(2, 1, 1.1, 2.2, 1, 0, 0.3, 1.1, 2.2, 1.1, 1.2, 2.5, 1.1, 0.3,
        0.5, 1.2, 2.5, 1.2, 1.3, 2.7, 1.2, 0.5, 0.6, 1.3),
  y = c(-0.5, 0, 1, 0.5, 0, 0.5, 1.5, 1, 0.5, 1, 2.1, 1.7, 1, 1.5,
        2.2, 2.1, 1.7, 2.1, 3.2, 2.8, 2.1, 2.2, 3.3, 3.2)
)

datapoly <- merge(values, positions, by = c("id"))
ggplot(datapoly, aes(x = x, y = y)) +
  geom_polygon(aes(group = id, fill = id)) +
  scale_fill_biolizard(type = "discrete", scheme = "sequential") +
  lizard_style()


```
## Simple box plot

```{r}
ggplot(data = mtcars, aes(x = as_factor(gear), y = mpg)) +
  geom_boxplot() +
  labs(
    title = "Miles per Gallon vs. Gears",
    x = "Gears",
    y = "Miles per Gallon"
  ) +
  lizard_style()

ggplot(data = mtcars, aes(x = as_factor(gear), y = mpg)) +
  geom_violin() +
  labs(
    title = "Miles per Gallon vs. Gears",
    x = "Gears",
    y = "Miles per Gallon"
  ) +
  lizard_style()
```

## Simple density plot

```{r}
ggplot(data = mtcars, aes(x = mpg)) +
  geom_density() +
  lizard_style()
```

## Simple scatterplot

```{r}

ggplot(data = mtcars, aes(x = hp, y = mpg)) +
  geom_point() +
  geom_smooth() +
  labs(
    title = "Miles per Gallon vs. horsepower",
    x = "Horsepower",
    y = "Miles per Gallon",
    color = "Miles per Gallon",
    fill = "Miles per Gallon"
  )

```
## Chicago plots

Idea from  ["A ggplot2 Tutorial for Beautiful Plotting in R"](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/).

```{r}
if (!requireNamespace("readr", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("readr")
}
library(ggplot2)
chic <- readr::read_csv("https://figshare.com/ndownloader/files/42307179") #An example dataset

g <- ggplot(chic, aes(x = season, y = o3,
                   color = season)) +
    labs(x = "Season", y = "Ozone") + 
    geom_violin(fill = "gray80", linewidth = 1, alpha = .5) +
    geom_jitter(alpha = .25, width = .3) +
    coord_flip()
g

g + lizard_style() +
  scale_color_biolizard()
```
```{r}
if (!requireNamespace(c("corrr","forcats"), quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages(c("corrr","forcats"))
}
library(corrr)
library(forcats)

corm <-
  chic |>
  dplyr::select(temp, dewpoint, pm10, o3) |>
  corrr::correlate(diagonal = 1) |>
  corrr::shave(upper = FALSE)

corm <- corm |>
  tidyr::pivot_longer(
    cols = -term,
    names_to = "colname",
    values_to = "corr"
  ) |>
  dplyr::mutate(
    rowname = forcats::fct_inorder(term),
    colname = forcats::fct_inorder(colname),
    label = dplyr::if_else(is.na(corr), "", sprintf("%1.2f", corr))
  )

g <- ggplot(corm, aes(rowname, fct_rev(colname),
                     fill = corr)) +
      geom_tile() +
      geom_text(aes(label = label)) +
      coord_fixed(expand = FALSE) +
      labs(x = NULL, y = NULL) 
g

g + scale_fill_biolizard(type='continuous',scheme='divergent',limits = c(-1, 1),na.value = "white", name="Pearson\nCorrelation") +
        lizard_style() +
        theme(legend.position = c(.95, .8))

```

```{r}
g <- ggplot(chic, aes(temp, o3)) +
        geom_hex(color = "grey") +
        labs(x = "Temperature (°F)", y = "Ozone Level", title= 'How Temperature Affects Ozone Levels',
             subtitle='Hexagonal Binning of Ozone Levels vs. Temperature in °F')
        
g

g <- g + lizard_style() +
      scale_fill_biolizard(type='continuous',scheme='sequential')
g

finalise_lizardplot(g,source="Source: The Chicago dataset (https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/)")
```

## Maps

note: to install 'sf' on ubuntu I had to install libudunits2-dev and libgdal-dev:

`sudo apt-get update && sudo apt-get install libudunits2-dev libgdal-dev`

For rpm install 'udunits2-devel' (Fedora, EPEL, ...) or for brew install 'udunits' (OSX)

+ libgdal-dev (TODO: find right rpm and brew package..)

```{r}
if (!requireNamespace("sf", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("sf")
}
if (!requireNamespace("rnaturalearth", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("rnaturalearth")
}
if (!requireNamespace("rnaturalearthdata", quietly = TRUE)) {
  # Install the package if it's not installed
  install.packages("rnaturalearthdata")
}

library("sf")
library("rnaturalearth")
library("rnaturalearthdata")

world <- ne_countries(scale = "medium", returnclass = "sf")

BLZ_offices <- c("Belgium", "Netherlands", "Switzerland", "United States of America")
world$BLZ_office <- ifelse(world$name_en %in% BLZ_offices, "yes", "no")
world$BLZ_office <- factor(world$BLZ_office, levels = c("yes", "no"))

ggplot(data = world, aes(fill = BLZ_office)) +
  geom_sf() +
  scale_fill_biolizard(type = "discrete", scheme = "qualitative") +
  lizard_style()

# coloring the ocean, as in the map on our website
ggplot(data = world, aes(fill = BLZ_office)) +
  geom_sf() +
  scale_fill_manual(values = c("#e9b940", "#01a086")) +
  lizard_style() +
  theme(panel.background = element_rect(fill = "#1e2237"),
        panel.grid = element_blank())

# europe only
eu_plot <- ggplot(data = dplyr::filter(world, continent == "Europe"), aes(fill = BLZ_office)) +
  geom_sf() +
  scale_fill_manual(values = c("#e9b940", "#01a086")) +
  lizard_style() +
  theme(panel.background = element_rect(fill = "#1e2237"),
        panel.grid = element_blank()) +
  coord_sf(xlim = c(-25,50), ylim = c(35,70), expand = FALSE)
eu_plot


finalise_lizardplot(eu_plot,source="BioLizard offices in Europe")
```
